### 2019-11-01 submission/response_to_reviewers.md

---
output:
  pdf_document: default
  html_document: default
---
### Editor Comments:

**1. Both reviewers commented about the appropriateness of strain that was used in your experiments. Please provide a justification for why this strain was chosen.**

> We have responded to the reviewers' comments below regarding why we chose to test *C. difficile* 630 and have added the following sentence to the discussion to acknowledge the strain choice as a potential limitation: 
"The type of *C. difficile* strain type could also be an important contributing factor, our study was limited in that we only tested *C. difficile* 630 (ribotype 012)."

**2. Both reviewers commented on the low number of mice used in the experiments, perhaps making it hard to pick up any small differences. Please add some additional discussion around this point.**

> We have responded to the reviewers' comments below and added the following sentence to the discussion regarding the low number of mice used:
"One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in specific bacteria genera."

**3. Lastly, from my own reading of the paper, I recall that rodent gut microflora consist mainly of Gram-positive organisms, whereas human gut flora is mainly Gram-negatives and anaerobes. One would presume that any treatment that disrupts the microbiome would interact very differently with these two environments, and that the rodent microbiome may not accurately reflect what happens in the human gut. This caveat is hinted at in lines 106-107 of your manuscript, but I would like you to make this point more clearly.**

> We have added an additional sentence to precede the sentence from lines 106-107 to draw attention to the differences in intestinal microbiome composition between mice and humans: 
"*While the intestinal microbiomes of both humans and mice are dominated by the Bacteroidetes and Firmicutes phyla, there are significant differences in the relative abundances of genera that are present and some genera are unique to each mammal (Hugenholtz & de Vos Cell. Mol. Life Sci. 2018), differences that may partly explain our results.* The microbiota and physiological differences between humans and mice may limit the usefulness of employing mouse models to study the impact of PPIs on the microbiota and CDIs."

### Reviewer #1 (Comments for the Author):

**In their manuscript Tomkovich et al. address the question whether the correlative association of proton pump inhibitors with the occurence of CDI is due to the changes in the microbiome following PPI administration as it is speculated in the field. This correlation drawn mainly from epidemiological studies and typically retrospectively, lacks experimental evaluations. Hence, the authors used the mouse colonization model of C. difficile. They checked the changes in the microbiome and the colonization efficiencies in mice treated either with the PPI, omeprazole or clindamycin or with a commbination of both. They show that efficient colonization is dependent on clindamycin treatment prior to infction and that omeprazole alone has no significant effect on the diversity of the microbiome, and does not result in effective colonzation by C. difficile. As this is the first attempt to experimentally address a question arising from medical practice and observation, the study and its findings are interesting.**
**Neverteheless, the study has some points that need to be addressed at least in the discussion before publication**

**1. Is it correct that the study was perfomed on max. 4 mice/treatment group? This might be too low a number, especially in case of the colonization studies to draw conclusions.**

> The reviewer is correct that the number of mice per group was low. We state the exact number of mice per treatment group in Figure 1A (5 mice each for 2 of the treatment groups and 4 mice for the 3rd treatment group). We agree that this is a low number, but would argue that the low microbiome variation across laboratory mice means less mice are needed compared to human epidemiological studies when examining *C. difficile* colonization. Previous *C. difficile* infection mouse model studies from our group and a close collaborator have used 3-5 mice per treatment group (see Koenigsknecht et al. Infection and Immunity 2015 and Theriot et al. Gut Microbes 2011 ) and were still able to draw meaningful conclusions. We have added the following sentence to the discussion to acknowledge the low number of mice used for these experiments as a limitation:
"One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in specific bacteria genera."

**2. The authors should address the fact that they have used only C. difficile str. 630 for their experiments. This is a ribotype 012 strain that is not predominant in the clinical setting. Which ribotypes are typically observed in PPI associated CDI, is there a tendency towards a certain ribotype? Could this have an impact on the interpretation of the data derived from the model described in this manuscript?**

> We thank the reviewers for pointing out that the type of *C. difficile* strain could be an important factor influencing the association between PPIs and CDIs observed in human epidemiological studies. Unfortunately, the studies to date seldom examine the *C. difficile* strains found in PPI users that develop CDIs. The studies that do discuss *C. difficile* strains merely note the type of *C. difficile* frequently found in the region that the study was conducted (Roughead et al. Expert Opinion on Drug Safety 2016; McDonald et al. JAMA Internal Medicine 2015; Chitnis et al. JAMA Internal Medicine 2013; Vardakas et al. International Journal of Infections Diseases 2012; Dalton et al. Alimentary Pharmacology & Therapeutics 2008). In addition to our lab's prior experience with the *C. difficile* 630 infection mouse model, we also decided to use *C. difficile* 630 because of a recent study that demonstrated three proton pump inhibitors had no effect on *in vitro C. difficile* 630 growth (Maier et al. Nature 2018). We have added the following sentence to the discussion to acknowledge that the type of *C. difficile* strain is another potential contributing factor to the association between PPIs and CDIs seen in human epidemiological studies:
"The type of *C. difficile* strain type could also be an important contributing factor, however our study was limited in that we only tested *C. difficile* 630 (ribotype 012)."

**3. As the manuscript is pretty concise and has in total 4 figures, the supplementary figures can be integrated into the main body.**

> We have submitted this manuscript as an observation, which is limited to 2 figures total. Additionally, in response to Reviewer #2, we have expanded the size of all the panels in our figures, limiting our ability to include additional panels for the 2 main text figures. Because of the changes to the sizes of the figures, we have opted to keep the supplementary figures as supplemental.

**5. There are some studies that looked into the effect of PPIs on microbiomes in cats and infants, which seem to be in accordance with the findings in this manuscript. Hence, the authors should include these in their discussion:**
**- Schmid SM, et al.,2018**
**- Castellani C, et al., 2017**

> Thank you for the suggestion, we have added the following sentence to the discussion referring to these studies to indicate that other groups have also observed minimal changes in the mammalian microbiome after proton pump inhibitor treatment:
"Interestingly, other groups examining fecal microbiota communities before and after PPI administration to healthy cats and infants with gastroesophageal reflux disease, found PPIs have minimal effects on fecal bacterial community structures, although there were a few significant changes in specific genera (Schmid et al. Frontiers in Veterinary Science 2018; Castellani et al. Frontiers in Cellular and Infection Microbiology 2017)."

**6. It is becoming more and more accepted that Clostridium difficile, due to its biology, is a separate genus, resulting in Clostridioides difficile as the accepted name. The authors should give credit to this and at least state it as Clostridium (Clostridioides) difficile in the text the first time they use it.**

> We appreciate the suggestion and have added "*Clostridium* (*Clostridioides*) *difficile*" to the abstract, importance, and introduction sections of our manuscript.

### Reviewer #2 (Comments for the Author):

**The manuscript by Tomkovich et al tests if the PPI omeprazole can influence the fecal microbiome of mice and promote C. difficile infection. The study is based on epidemiological findings in humans that associate increased PPI use with CDI as well as a few mouse studies that implicate PPIs in CDI. The data presented here show no significant impact of omeprazole on the microbiome of mice after 7 days of treatment and the mice did not become susceptible to C. diff colonization. This is a small study but adds to the literature in that it conclusively shows no impact of omeprazole on the microbiome.**

**1. The size of the figures and the fonts are very small and should be improved in a resubmission. It makes it difficult to see individual points in the graphs.**

> We have adjusted the sizes of all figures and increased the sizing of the fonts and points in the graphs.

**2. The choice of CD630 and dose may have a big impact on the results and comparison with previous studies. In ref. 15 the authors used a more virulent strain (VPi) and at a 10,000 fold increased concentration. The authors do a good job of citing other factors that may influence the results (like the increased pH of the mouse stomach) but this should also be noted for readers.**

> We thank the reviewer for pointing out *C. difficile* strain type as an additional factor that could explain the association between proton pump inhibitors and CDIs in human studies, which was also pointed out by Reviewer # 1. In response, we have added the following sentence to the discussion:
"The type of *C. difficile* strain type could also be an important contributing factor, however our study was limited in that we only tested *C. difficile* 630 (ribotype 012)."

**3. Please provide evidence for how the dose of 40mg/kg of OMZ corresponds to what humans get in a dose.**

>  Omeprazole is prescribed to adults in 20 mg or 40 mg doses that are taken once per day. The 40 mg/kg dose administered to the mice in our study is higher than what is administered to humans translating to ~ 3.25 mg/kg (compared to ~ 0.33-0.66 mg/kg for the prescribed human dose) based on a reference body weight of 60 kg for humans and 0.02 kg for mice (converted according to the calculation described in Nair & Jacob J Basic Clin Pharm. 2016). However, previous studies have justified administering omeprazole at higher doses to mice because the drug administered to the mice lacks an enteric coating (Llorente et al. Nature Communications 2017) and is eliminated more quickly compared to humans (plasma elimination half life of 5-15 minutes compared to 1 hour for humans, see Regardh et al. Scandinavian Journal of Gastroenterology 1985). The 40 mg/kg dose chosen for our study is in line with what was used by one of the previous studies that examined the proton pump inhibitor esomeprazole and *C. difficile* in mice (see Hung et al. Journal of Infectious Diseases 2015). We have added the following sentence to the methods section to account for the higher dosing:
"Although the omeprazole dose administered to mice is higher than the recommended dose for humans, omeprazole has a shorter half-life in mice compared to humans (Regardh et al. Scandinavian Journal of Gastroenterology 1985) and lacks an enteric coating (Llorente et al. Nature Communications 2017)."

**4. Although OMZ did not have a major impact on community structure by itself, in figure 2 it seems that it did have an impact on a few groups that were treated with clindamycin. Although the P-values aren't significant, this could be mainly due to the low number of mice that were used in the study as a few of the genera showed 10-fold changes. I can see why it isn't commented on due to the lack of stat significance, but it is pretty striking as presented in the figure.**

> We agree with the reviewer regarding Figure 2C, where the relative abundances for specific taxa in the Clindamycin + omeprazole treated mice were typically in between the clindamycin-treated and omeprazole-treated mice. Beyond the P-values not being significant, we decided not to comment on this because we suspect that the majority of the differences in bacterial relative abundances across groups is driven by the clindamycin treatment. We have added the following sentence to the discussion to acknowledge the low number of mice used as a potential limitation in detecting omeprazole-induced changes in the fecal microbiota:
"One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in bacteria genera."

**5. The final sentence of the importance section is a run-on sentence and needs to be broken up or simplified.**

> To simplify the sentence, we have removed the part that elaborates on what other factors ("composition of the starting bacterial community, comorbidities, and use of additional medications") might be contributing to the correlations between PPIs and CDIs seen in humans.

**6. There are many typos and missed italics in the references.**

> Thank you for pointing these errors out, typos have been corrected and italics have been added to the references.

### 2019-11-01 submission/manuscript.Rmd
Get rid of warnings about packages being built under specific R versions when markdown is knitted
Respond to reviewers' comments

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

@@

## Abstract
Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium difficile* infections (CDIs) in humans. We assessed how PPI treatment alters the fecal microbiota and whether treatment promotes CDIs in a mouse model. Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment phase, the day of *C. difficile* challenge, and the following 9 days. We found that mice treated with omeprazole were not colonized by *C. difficile*. When omeprazole treatment was combined with a single clindamycin treatment, one cage of mice remained resistant to *C. difficile* colonization, while the other cage was colonized. Treating mice with only clindamycin followed by challenge resulted in *C. difficile* colonization. 16S rRNA gene sequencing analysis revealed that omeprazole had minimal impact on the structure of the murine microbiota throughout the 16 days of omeprazole exposure. These results suggest omeprazole treatment alone is not sufficient to disrupt microbiota resistance to *C. difficile* infection in mice that are normally resistant in the absence of antibiotic treatment.

## Importance
Antibiotics are the primary risk factor for *Clostridium difficile* infections (CDIs), but other factors may also increase a person's risk. In epidemiological studies, proton pump inhibitor (PPI) use has been associated with CDI incidence and recurrence. PPIs have also been associated with alterations in the human intestinal microbiota in observational and interventional studies. We evaluated the effects of the PPI omeprazole on the structure of the murine intestinal microbiota and its ability to disrupt colonization resistance to *C. difficile*. We found omeprazole treatment had minimal impact on the murine fecal microbiota and did not promote *C. difficile* colonization. Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs seen in humans or whether aspects of murine physiology may limit its utility to test these types of hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium difficile* infections (CDIs) [@Schubert2015]. It is less clear whether other human medications that impact the microbiota also influence *C. difficile* colonization resistance. Multiple epidemiological studies have suggested an association between proton pump inhibitor (PPI) use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019]. There have also been a number of large cohort studies and interventional clinical trials that demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018]. PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018]. In human fecal samples, PPI use results in increases in *Enterococcaceae*, *Lactobacillaceae*, *Micrococcaceae*, *Staphylococcaceae* and *Streptococcaceae* and decreases in *Ruminococcaceae* [@Imhann2015; @Imhann2017; @Freedberg2015; @maier2018extensive]. Several of these taxa have also been associated with *C. difficile* colonization in humans [@Schuberte01021-14].      
Unfortunately, the studies suggesting a link between PPIs and *C. difficile* were retrospective and did not evaluate changes in the microbiome [@tariq2017association; @nehra2018proton; @Elias2019]. Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use explain the association between PPIs and CDIs. Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].  Here, we evaluated the impact of daily PPI treatment with omeprazole on the murine microbiome and susceptibility to *C. difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C. difficile* to colonize but is mild enough that *C. difficile* is cleared within 10 days [@Jenior2018].

@@

**Conclusions.** The PPI omeprazole did not meaningfully impact the structure of the gut microbiota and did not promote *C. difficile* infection in mice. Our findings that omeprazole treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated the PPI lansoprazole had more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018]. The same group demonstrated lansoprazole treatment increased the stomach pH in mice [@Yasutomi2018], which may improve survival of bacteria passing through the stomach. We did not find significant changes in the relative abundances of the taxa observed to be significantly impacted by PPI use in human studies. However, 3 of the human-associated taxa were absent or at low abundance in our mice. Although our fecal microbiota findings are comparable to what has been shown in another mouse study [@Yasutomi2018], whether PPI-induced changes in specific bacterial abundances observed in humans play a role in CDIs remains to be determined.    

@@

Our study also extended previous work examining PPIs and *C. difficile* in mice by incorporating the contribution of the intestinal microbiota. We found omeprazole had no significant impact on bacterial taxa within the murine intestinal microbiota over the 16-day experiment. In contrast to previous work with PPIs [@Kaur2007; @kaur2011effect; @hung2015proton], omeprazole did not alter *C. difficile* colonization resistance in mice. 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony. These two genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to our observation that PPI treatment had no effect on *C. difficile* colonization in our CDI mouse model. Gastrointestinal physiological differences, particularly the higher stomach pH in mice (pH 3-4) compared to humans (pH 1) [@Hugenholtz2017] could also explain why omeprazole had a limited impact on the murine microbiome. The microbiota and physiological differences between humans and mice may limit the usefulness of employing mouse models to study the impact of PPIs on the microbiota and CDIs.   
Beyond microbiome differences, factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence. This study addressed the impact of PPIs with or without antibiotics on a murine model of CDI, and found PPIs did not promote *C. difficile* colonization. The epidemiological evidence linking PPIs to CDIs is primarily from observational studies, which makes determining causality and whether other risk factors play a role challenging [@Eze2017]. Future studies are needed to determine whether age, other comorbidities and bacterial strains that are less common in mice can increase the risk of CDIs or recurrent CDIs when combined with PPI treatment.

@@

**Drug treatments.**
Omeprazole (Sigma Aldrich) was prepared in a vehicle solution of 40% polyethylene glycol 400 (Sigma-Aldrich) in phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen aliquots and diluted to an 8 mg/mL prior to gavage. All mice received 40 mg/kg omeprazole (a dose previously used in mouse experiments [@hung2015proton]) or vehicle solution once per day through the duration of the experiment with treatment starting 7 days before *C. difficile* challenge (Figure 1A). One day prior to *C. difficile* challenge, 2 groups of mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle [@Jenior2018]. All drugs were filter sterilized through a 0.22 micron syringe filter before administration to animals.

>>>>

```{r knitr_settings, tidy=TRUE, eval=TRUE, warning = FALSE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

@@

## Abstract
Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) in humans. We assessed how PPI treatment alters the fecal microbiota and whether treatment promotes CDIs in a mouse model. Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment phase, the day of *C. difficile* challenge, and the following 9 days. We found that mice treated with omeprazole were not colonized by *C. difficile*. When omeprazole treatment was combined with a single clindamycin treatment, one cage of mice remained resistant to *C. difficile* colonization, while the other cage was colonized. Treating mice with only clindamycin followed by challenge resulted in *C. difficile* colonization. 16S rRNA gene sequencing analysis revealed that omeprazole had minimal impact on the structure of the murine microbiota throughout the 16 days of omeprazole exposure. These results suggest omeprazole treatment alone is not sufficient to disrupt microbiota resistance to *C. difficile* infection in mice that are normally resistant in the absence of antibiotic treatment.

## Importance
Antibiotics are the primary risk factor for *Clostridium* (*Clostridioides*) *difficile* infections (CDIs), but other factors may also increase a person's risk. In epidemiological studies, proton pump inhibitor (PPI) use has been associated with CDI incidence and recurrence. PPIs have also been associated with alterations in the human intestinal microbiota in observational and interventional studies. We evaluated the effects of the PPI omeprazole on the structure of the murine intestinal microbiota and its ability to disrupt colonization resistance to *C. difficile*. We found omeprazole treatment had minimal impact on the murine fecal microbiota and did not promote *C. difficile* colonization. Further studies are needed to determine whether other factors contribute to the association between PPIs and CDIs seen in humans or whether aspects of murine physiology may limit its utility to test these types of hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) [@Schubert2015]. It is less clear whether other human medications that impact the microbiota also influence *C. difficile* colonization resistance. Multiple epidemiological studies have suggested an association between proton pump inhibitor (PPI) use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019]. There have also been a number of large cohort studies and interventional clinical trials that demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018]. PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018]. In human fecal samples, PPI use results in increases in *Enterococcaceae*, *Lactobacillaceae*, *Micrococcaceae*, *Staphylococcaceae* and *Streptococcaceae* and decreases in *Ruminococcaceae* [@Imhann2015; @Imhann2017; @Freedberg2015; @maier2018extensive]. Several of these taxa have also been associated with *C. difficile* colonization in humans [@Schuberte01021-14].      
Unfortunately, the studies suggesting a link between PPIs and *C. difficile* were retrospective and did not evaluate changes in the microbiome [@tariq2017association; @nehra2018proton; @Elias2019]. Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use explain the association between PPIs and CDIs. Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].  Here, we evaluated the impact of daily PPI treatment with omeprazole on the murine microbiome and susceptibility to *C. difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C. difficile* to colonize but is mild enough that *C. difficile* is cleared within 10 days [@Jenior2018].

@@

**Conclusions.** The PPI omeprazole did not meaningfully impact the structure of the gut microbiota and did not promote *C. difficile* infection in mice. Our findings that omeprazole treatment had minimal impact on the fecal microbiome were comparable to another PPI mouse study that indicated the PPI lansoprazole had more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018]. The same group demonstrated lansoprazole treatment increased the stomach pH in mice [@Yasutomi2018], which may improve survival of bacteria passing through the stomach. We did not find significant changes in the relative abundances of the taxa observed to be significantly impacted by PPI use in human studies. However, 3 of the human-associated taxa were absent or at low abundance in our mice. Interestingly, other groups examining fecal microbiota communities before and after PPI administration to healthy cats and infants with gastroesophageal reflux disease, found PPIs have minimal effects on fecal bacterial community structures, although there were a few significant changes in specific genera [@Schmid2018; @Castellani2017]. One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in specific bacteria genera. Although our fecal microbiota findings are comparable to what has been shown in another mouse study [@Yasutomi2018], whether PPI-induced changes in specific bacterial abundances observed in humans play a role in CDIs remains to be determined.    

@@

Our study also extended previous work examining PPIs and *C. difficile* in mice by incorporating the contribution of the intestinal microbiota. We found omeprazole had no significant impact on bacterial taxa within the murine intestinal microbiota over the 16-day experiment. In contrast to previous work with PPIs [@Kaur2007; @kaur2011effect; @hung2015proton], omeprazole did not alter *C. difficile* colonization resistance in mice. 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare genera in our C57BL/6 mouse colony. These two genera could be important contributors to the associations between PPIs and CDIs in humans, and could be a contributing factor to our observation that PPI treatment had no effect on *C. difficile* colonization in our CDI mouse model. While the intestinal microbiomes of both humans and mice are dominated by the *Bacteroidetes* and *Firmicutes* phyla, there are significant differences in the relative abundances of genera that are present and some genera are unique to each mammal [@Hugenholtz2018], differences that may partly explain our results. Gastrointestinal physiological differences, particularly the higher stomach pH in mice (pH 3-4) compared to humans (pH 1) [@Hugenholtz2018] could also explain why omeprazole had a limited impact on the murine microbiome. The microbiota and physiological differences between humans and mice may limit the usefulness of employing mouse models to study the impact of PPIs on the microbiota and CDIs.   
Beyond microbiome differences, factors such as age, body mass index, comorbidities, and use of other medications in human studies may also be contributing to the association between PPIs and CDI incidence or recurrence. The type of *C. difficile* strain type could also be an important contributing factor, however our study was limited in that we only tested *C. difficile* 630 (ribotype 012). This study addressed the impact of PPIs with or without antibiotics on a murine model of CDI, and found PPIs did not promote *C. difficile* colonization. The epidemiological evidence linking PPIs to CDIs is primarily from observational studies, which makes determining causality and whether other risk factors play a role challenging [@Eze2017]. Future studies are needed to determine whether age, other comorbidities and bacterial strains that are less common in mice can increase the risk of CDIs or recurrent CDIs when combined with PPI treatment.

@@

**Drug treatments.**
Omeprazole (Sigma Aldrich) was prepared in a vehicle solution of 40% polyethylene glycol 400 (Sigma-Aldrich) in phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen aliquots and diluted to an 8 mg/mL prior to gavage. All mice received 40 mg/kg omeprazole (a dose previously used in mouse experiments [@hung2015proton]) or vehicle solution once per day through the duration of the experiment with treatment starting 7 days before *C. difficile* challenge (Figure 1A). Although the omeprazole dose administered to mice is higher than the recommended dose for humans, omeprazole has a shorter half-life in mice compared to humans [@Regrdh1985] and lacks an enteric coating [@Llorente2017]. One day prior to *C. difficile* challenge, 2 groups of mice received an intraperitoneal injection of 10 mg/kg clindamycin or sterile saline vehicle [@Jenior2018]. All drugs were filter sterilized through a 0.22 micron syringe filter before administration to animals.
